An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-03-17 paper.tex
Fixed tree space
Minor changes to IV.

We obtain $10^4$ individual model MCMC samples within each model, and
use the kD-tree interpolation method described above to propose
inter-model jumps in an RJMCMC.  The acceptance rate of inter-model
jumps is approximately 0.8.  To explore how the efficiency of the
method degrades as the interpolation becomes less accurate, we
artificially truncated the kD tree with higher and higher numbers of
points in each box (this can be accomplished during the neighborhood
search phase by stopping the search for a box when one is found
containing the desired number of points).  For each truncation, we
performed an RJMCMC with the resulting interpolated jump proposal.
The acceptance rate versus truncation number appear in Figure
\ref{fig:acceptRate}.  The more points in each box of the tree when
the search is truncated, the lower the acceptance probability; when
points are drawn from the top level of the tree, the acceptance
probability asymptotes to the naive draw from the prior ($\sim 5\%$).

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate
    versus the number of points per box when the kD-tree neighborhood
    search is truncated.  As the number of points per box increases,
    and the interpolation becomes less accurate, the acceptance rate
    falls, asymptoting to the rate for naive draws from the uniform
    prior (about $5\%$).}
\end{figure}

The relative error on the determination of the evidence ratio scales
with $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of inter-model
transitions in the RJMCMC.  Thus, as the acceptance rate of
inter-model jumps goes down, the RJMCMC must run longer to achieve a
desired accuracy in the evidence ratio.  By boosting the acceptance
rate of inter-model jumps significantly, the interpolation method
described above can dramatically improve the runtime of an RJMCMC.

>>>>

We obtain $10^4$ single-model MCMC samples by independently running MCMC within each model, and use the kD-tree interpolation method described above to propose
inter-model jumps in an RJMCMC.  The acceptance rate of inter-model
jumps is approximately 0.8.  To explore how the efficiency of the
method degrades as the interpolation becomes less accurate, we
artificially truncated the kD tree with higher and higher numbers of
points in each box (this can be accomplished during the neighborhood
search phase by stopping the search for a box when one is found
containing the desired number of points).  For each truncation choice, we
performed an RJMCMC with the resulting interpolated jump proposal.
The acceptance rate is plotted against the number of single-model MCMC points per box (kD-tree leaf) in Figure \ref{fig:acceptRate}.  The more points in each leaf of the tree when
the search is truncated, the lower the acceptance probability; when
points are drawn from the top level of the tree, the acceptance
probability asymptotes to the naive draw from the prior ($\sim 5\%$).

\begin{figure}
  \begin{center}
    \includegraphics[width=0.8\columnwidth]{acceptRate}
  \end{center}
  \caption{\label{fig:acceptRate} The inter-model acceptance rate
    versus the number of points per box when the kD-tree neighborhood
    search is truncated.  As the number of points per box increases,
    and the interpolation becomes less accurate, the acceptance rate
    falls, asymptoting to the rate for naive draws from the uniform
    prior (about $5\%$ for this data set).}
\end{figure}

The relative error on the determination of the Bayes factor (evidence ratio) scales
with $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of inter-model
transitions in the RJMCMC.  Thus, as the acceptance rate of
inter-model jumps goes down, the RJMCMC must run longer to achieve a
desired accuracy in the Bayes factor.  By boosting the acceptance
rate of inter-model jumps significantly, the interpolation method
described above can dramatically improve the runtime of an RJMCMC.
